---
title: "R Markdown First Assignment"
author: "Mirella Castaneda"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
# Import libraries or install packages

#install.packages('tidyverse')
library(tidyverse)
library(knitr)

#Installing other packages

#install.packages("kableExtra")
library('kableExtra')

library(RColorBrewer)

#install.packages('tableone')
library('tableone')

library('table1')

```



```{r}
# Read in data

mydf <- read_csv(file = file.choose())

# Full view of data
print(mydf, n=150)

# Quick view 
mydf
```

```{r graph}

# A graph showing the IQ levels by location status

t <- ggplot(mydf)
  

t + geom_col(aes(Smelter, IQ)) +
  labs(x = "Location Status", y = 'IQ Levels', 
       title = "IQ Levels by Location Status")

```



```{r}
# At least one nicely formatted table (hint: try the 'kable' function in the knitr R package or one of the popular table R packages like tableone, table1, or gtsummary)

# Used table1 package

label(mydf$IQ) <- 'IQ Levels'
label(mydf$Smelter) <- 'Locations Status'
table1(~ IQ | Smelter, data = mydf)

```


### The bar chart shows the total summed IQ levels categorized by the proximity to the smelter (far vs near), where IQ levels appear overall lower closer to the smelter. Table 1 details the summary statistics for each group across the sample of 124 participants. Participants living farther from the smelter had a higher mean IQ (mean=106,SD=112) compared to those living closer(near) the smelter (mean=89.2,SD=12.2).






```{r kable, results='asis'}
# Using knit R package (for practice)


far_mean <- round(mean((mydf$IQ[mydf$Smelter =="Far"])),2)
far_sd <- round(sd((mydf$IQ[mydf$Smelter =="Far"])),2)
near_mean <- round(mean((mydf$IQ[mydf$Smelter =="Near"])),2)
near_sd <- round(sd((mydf$IQ[mydf$Smelter =="Near"])),2)

table_list <- list(far_mean,far_sd,near_mean,near_sd)

kable(table_list)

  
```

```{r}

# Calculate summary statistics by Smelter group
iq_summary <- mydf %>%
  group_by(Smelter) %>%
  summarize(
    N = n(),
    `Mean IQ` = round(mean(IQ, na.rm = TRUE), 1),
    `SD IQ`   = round(sd(IQ, na.rm = TRUE), 1),
    `Median IQ` = round(median(IQ, na.rm = TRUE), 1)
  )

# Output as a formatted table
kable(iq_summary, caption = "Summary Statistics of IQ Levels by Location Status", align = "lccrr")

```




<!-- In-line calculations giving the values of the means, that will be updated if the data are changed (hint: use syntax `r `) -->

Participants living farther from the smelter had a higher mean IQ 
(Mean = `r far_mean`, SD = `r far_sd`) 
compared to those living closer 
(Mean = `r near_mean`, SD = `r near_sd`).

